# Study dataset

In this tutorial, I still use the pulic smoking study dataset from NCBI
Gene Expression Omnibus (GEO) [database](https://www.ncbi.nlm.nih.gov/gds) - GDS3713 (GSE18723).

## Extracting data from NCBI Gene Expression Omnibus databse

Using `GEOquery` library to retrieve data from NCBI database. The GDS dataset
is stored in the `gds` R4 class object.

```{r , cache=TRUE}
library(here)
library(GEOquery)
library(dplyr)
gds <- getGEO("GDS3713")
# gds@header # meta data of the experiment
# gds@dataTable@table # expression data
# gds@dataTable@columns # phenotype data
```

structure of the `gds` object

```{r }
str(gds, max.level = 3)
```

### Summary of the study
This study aim to analyze peripheral circulating B cells from smoking and non-smoking healthy US white females. B cells are directly associated with the onset and development of many smoking-induced diseases. Results provide insight into the molecular basis of B cell involvement in smoking-related pathogenesis. It contains 39 smoker and 40 non-smokers.

### Assigning the expression and phenotype data
The expression data is stored in the `gds@dataTable@tabl` with gene in
The rows and subjects in the columns.
The phenotype data is store at `gds@dataTable@columns`.

```{r }
expr <- Table(gds)
summary(expr[, 1:4])
pheno <- Columns(gds)
summary(pheno)
```

## convert to GDS to expression set
Converting the gds object to expression set will generate a series of sub documents, including an annotation file of the genes in the expression data.

```{r }
eset <- GDS2eSet(gds)
dim(pData(eset)) # pheno data, data.frame
str(eset, max.level =4)
# gene annotations
annot_cols <- eset@featureData@varMetadata
annot_data <- eset@featureData@data
```
