---
title: Extracting data from NCBI Gene Expression Omnibus databse
subtitle: 'SAR: NA , Study: pathway analysis'
author:  Siying Huang (E0482362), Biomarker statistics team
date: 'created: 2023-06-01 , updated (`r Sys.Date()`)'
always_allow_html: true
output:
  html_document:
    css: sanofi.css
    code_folding: "show"
    toc: yes
    toc_float:
      collapse: no
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = T, comment = '',message = F, warning = F, error=F)
options(width = 100)

```
```{r libs}
library(here)
library(GEOquery)
library(dplyr)

# require data from NCBI Gene Expression Omnibus (GEO) [database](https://www.ncbi.nlm.nih.gov/gds)
# datasets:
# https://www.ncbi.nlm.nih.gov/sites/GDSbrowser?acc=GDS5022 AMI
# https://www.ncbi.nlm.nih.gov/sites/GDSbrowser?acc=GDS5074 AMI 48 hrs
# https://www.ncbi.nlm.nih.gov/sites/GDSbrowser?acc=GDS4500 AML train set
# https://www.ncbi.nlm.nih.gov/sites/GDSbrowser?acc=GDS4501 AML test set
# https://www.ncbi.nlm.nih.gov/sites/GDSbrowser?acc=GDS4181 AML
# https://www.ncbi.nlm.nih.gov/sites/GDSbrowser?acc=GDS4844 CF
# https://www.ncbi.nlm.nih.gov/sites/GDSbrowser?acc=GDS3115 CHF
# https://www.ncbi.nlm.nih.gov/sites/GDSbrowser?acc=GDS3383 chronic stress
# https://www.ncbi.nlm.nih.gov/sites/GDSbrowser?acc=GDS838 CML
# https://www.ncbi.nlm.nih.gov/sites/GDSbrowser?acc=GDS3503 exercise
# https://www.ncbi.nlm.nih.gov/sites/GDSbrowser?acc=GDS4345 GI cancer muscle tissue
# https://www.ncbi.nlm.nih.gov/sites/GDSbrowser?acc=GDS1412 hormone therapy
# https://www.ncbi.nlm.nih.gov/sites/GDSbrowser?acc=GDS3005 IL1 and IL6
# https://www.ncbi.nlm.nih.gov/sites/GDSbrowser?acc=GDS5260 Psoriasis
# https://www.ncbi.nlm.nih.gov/sites/GDSbrowser?acc=GDS3628 RA
# https://www.ncbi.nlm.nih.gov/sites/GDSbrowser?acc=GDS5403 RA and OA
# https://www.ncbi.nlm.nih.gov/sites/GDSbrowser?acc=GDS4266 renal transplant
# https://www.ncbi.nlm.nih.gov/sites/GDSbrowser?acc=GDS5186 RIA
# https://www.ncbi.nlm.nih.gov/sites/GDSbrowser?acc=GDS3898 social isolation
# https://www.ncbi.nlm.nih.gov/sites/GDSbrowser?acc=GDS4193 SLE
# https://www.ncbi.nlm.nih.gov/sites/GDSbrowser?acc=GDS3713 smoking
# https://www.ncbi.nlm.nih.gov/sites/GDSbrowser?acc=GDS1436 smoking
# https://www.ncbi.nlm.nih.gov/sites/GDSbrowser?acc=GDS1269 smoking
# https://www.ncbi.nlm.nih.gov/sites/GDSbrowser?acc=GDS534 smoking
# https://www.ncbi.nlm.nih.gov/sites/GDSbrowser?acc=GDS3318 scikle cell disease
# https://www.ncbi.nlm.nih.gov/sites/GDSbrowser?acc=GDS4719 SLE
# https://www.ncbi.nlm.nih.gov/sites/GDSbrowser?acc=GDS3920 MS
# https://www.ncbi.nlm.nih.gov/sites/GDSbrowser?acc=GDS3886 MS
# https://www.ncbi.nlm.nih.gov/sites/GDSbrowser?acc=GDS4270 UC
```

# Download data to the current working environment

Using `GEOquery` library to retrieve data from NCBI database. The GDS dataset
is stored in the `gds` R4 class object.

```{r }
gds <- getGEO("GDS3713")
# gds@header # meta data of the experiment
# gds@dataTable@table # expression data
# gds@dataTable@columns # phenotype data
```

structure of the gds object

```{r }
str(gds, max.level = 3)
```

## Summary of experiment

```{r }
show(Meta(gds))
```

## Assigning the expression and phenotype data
The expression data is stored in the `gds@dataTable@tabl` with gene in
The rows and subjects in the columns.
The phenotype data is store at `gds@dataTable@columns`.

```{r }
expr <- Table(gds)
summary(expr[, 1:4])
pheno <- Columns(gds)
summary(pheno)
```

# convert to GDS to expression set
Converting the gds object to expression set will generate a series of sub documents, including an annotation file of the genes in the expression data.

```{r }
eset <- GDS2eSet(gds)
dim(pData(eset)) # pheno data, data.frame
str(eset, max.level =4)
# gene annotations
annot_cols <- eset@featureData@varMetadata
annot_data <- eset@featureData@data
# experiment info
# eset@experimentData
# assay <- eset@assayData

```
```{r save, eval= F, echo = F}
saveRDS(expr, "data/expr.rds")
saveRDS(pheno, "data/pheno.rds")
saveRDS(annot_data, 'data/annot.rds')
```

<details><summary>Session Info</summary>

```{r }
sessionInfo()
```

</details>

```{r }
# Markdown --------------------------------------------------------
# rmarkdown::render('src/02_req_data.R', output_dir = 'output')
# knitr::spin('src/02_req_data.R', format = 'Rmd', knit = FALSE)
```

